---
title: "p8105-hw2-yw4662"
output: github_document
---

```{r setup}
library(tidyverse)
library(janitor)
library(haven)
```

## Problem 1

### 1. Clean the data in pols-month.csv

```{r}
pols <- read_csv("./data/pols-month.csv") %>%
  clean_names()%>%
  
  # Split mon into year, month, day as integers
  separate(mon, into = c("year", "month", "day"), sep = "-", convert = TRUE) %>%
  
  # Replace month number with month name
  mutate(month = month.name[month],
         # Create president from indicator cols
         president = if_else(prez_gop == 1, "gop", "dem"),
         president = factor(president, levels = c("gop", "dem"))) %>%
  
  # Remove prez_dem, prez_gop, and day variables
  select(-prez_dem, -prez_gop, -day)

pols
```

### 2. Clean the data in snp.csv

```{r}
snp <- read_csv("./data/snp.csv") %>%
  clean_names() %>%
  
  # Parse the data
  mutate(date = mdy(date),
         year = year(date),
         mon_n = month(date)) %>%
  mutate(month = month.name[mon_n]) %>%
  select(-mon_n, -date) %>%
  
  # Make year & month leading columns
  relocate(year, month, .before = 1) %>%
  
  # Arrange according to year and month
  arrange(year, month(match(month, month.name)))

snp
```

### 3. Tidy the unemployment data

```{r}
unemp <- read_csv("./data/unemployment.csv") %>%
  clean_names() %>%
  pivot_longer(
    cols = jan:dec,
    names_to = "month",
    values_to = "unemployment"
  ) %>%
  mutate(
    year = as.integer(year),
    month = str_to_title(month.name[match(month, tolower(month.abb))]) # matching
  ) %>%
  relocate(year, month, .before = 1) %>%
  arrange(year, match(month, month.name))

unemp
```

### 4. Merge the datasets

```{r}
mm <- factor(month.name, levels = month.name)

pols <- pols %>%
  mutate(
    year = as.integer(year),
    month = factor(month, levels = levels(mm))
  ) %>%
  arrange(year, month)

snp <- snp %>%
  mutate(
    year = as.integer(year),
    month = factor(month, levels = levels(mm))
  ) %>%
  arrange(year, month)

unemp <- unemp %>%
  mutate(
    year = as.integer(year),
    month = factor(month, levels = levels(mm))
  ) %>%
  arrange(year, month)

# Merging
merged <- pols %>%
  left_join(snp, by = c("year", "month")) %>%
  left_join(unemp, by = c("year", "month"))

merged
```

### 5. Comment

The **pols** table contains monthly U.S. politicians counts and a `president` indicator; it spans `r min(pols$year)` - `r max(pols$year)` with `r nrow(pols)` rows. The **snp** table adds monthly S&P returns, and the **unemp** table provides monthly unemployment rates for `r length(unique(unemp$year))` years.

After merging on `year` and `month`, the resulting dataset has `r nrow(merged)` rows × `r ncol(merged)` columns; key variables include `year`, `month`, `president`, `unemployment`, and the S&P return column from `close`.